Estimating Predictive Rate-Distortion Curves via Neural Variational Inference

### 2019-05-03 paper.tex
Updates from Overleaf

@@

This motivates the study of models that try to extract those key features of past observations that are most relevant to predicting the future.
%This results in a tradeoff between the cost of representing past observations and predicting the future well.
%This tradeoff has been formalized and computationally studied in a range of prior studies.
A general information-theoretic framework for this problem is provided by Predictive Rate-Distortion~~\citep{marzen-predictive-2016}, also known as the Past-Future Information Bottleneck~\citep{creutzig-past-future-2009}.
The Predictive Rate-Distortion tradeoff seeks to find an encoding of past observations that is maximally informative about future observations, while satisfying a bound on the amount of information that has to be stored.
More formally, this framework trades off prediction loss in the future, formalized as cross-entropy, with the cost of representing the past, formalized as the mutual information between the past observations and the compressed representations of the past.
Due to its information-theoretic nature, this framework is extremely general and applies to processes across vastly different domains.
It has been applied to linear dynamical systems \citep{creutzig-past-future-2009,amir2015past}, but is equally meaningful for discrete dynamical systems~\citep{marzen-predictive-2016}.
For biological systems that make predictions about their environment, this corresponds to placing an information-theoretic constraint on the computational cost used for conditioning actions on past observations~\citep{genewein2015bounded}.

@@

Optimal Causal Filtering (OCF, \citet{still-optimal-2010}) addresses the problem of estimating Predictive Rate-Distortion from a finite sample of observation trajectories.
It does so by constructing a matrix of observed frequencies of different pairs of past and future observations.
However, this method faces a series of challenges~\citep{marzen-predictive-2016}.
One is the curse of dimensionality:
Modeling long dependencies requires storing an exponential number of observations, which quickly becomes intractable for current computation methods.
This exponential growth is particularly problematic when dealing with processes with a large state space.
For instance, the number of distinct words in a human language as found in large-scale text data easily exceeds $10^5$, making storing and manipulating counts of longer word sequences very challenging.
A second challenge is that of overfitting:
When deploying a predictive model constructed via OCF on new data to predict upcoming observations, such a model can only succeed when the past sequences occurred in the sample that OCF was applied to.
This is because OCF relies on counts of full past and future observation sequences; it has no means to generalize to unseen past sequences.

@@

In Section~\ref{sec:background}, we formally introduce predictive rate-distortion, and discuss related notions of predictive complexity.
In Section~\ref{sec:oce}, we describe the prior method of Optimal Causel Filtering (OCF).
In Section~\ref{sec:nprd}, we describe our method NPRD.
In Section~\ref{sec:experiments}, we validate NPRD on processes whose predictive rate-distortion is analytically known, showing that it finds essentially optimal predictive models.
In Section~\ref{sec:language}, we apply NPRD to data from five languages, providing the first estimates of predictive rate-distortion of natural language.

@@

The main prior method for estimating predictive rate-distortion from data is Optimal Causal Filtering (OCF, \citet{still-optimal-2010}).
This method approximates predictive rate-distortion using two approximations:
First, it replaces semi-infinite pasts and futures with bounded-length contexts, i.e., pairs of finite past contexts $(\finitepast := X_{-M}\dots X_{-1})$ and future contexts $(\finitefuture := X_{0}\dots X_{M-1})$ of some finite length $M$.\footnote{It is not crucial that past and future contexts have the same lengths, and indeed \citet{still-optimal-2010} do not assume this. We do assume equal length throughout this paper for consistency with our experiments, though nothing depends on this.}
The PRD objective (\ref{eq:ib}) then becomes (\ref{eq:ib-oce}), aiming to predict length-$M$ finite futures from summary codes $Z$ of length-$M$ finite pasts:
\begin{equation}\label{eq:ib-oce}
	\max_{Z : Z \bot \finitefuture | \finitepast} \left( \operatorname{I}[Z, \finitefuture] - \lambda \cdot \operatorname{I}[\finitepast, Z] \right).
\end{equation}
Second, OCF estimates information measures directly from the observed counts of $(\finitepast), (\finitefuture)$ using the plug-in estimator of mutual information. %\rljf{Should be \dots instead of commas?}
With such an estimator, the problem in (\ref{eq:ib-oce}) can be solved using a variant of the Blahut-Arimoto algorithm \citep{tishby-information-1999}, obtaining an encoder $P(Z|\finitepast)$ that maps each observed past sequence $\finitepast$ to a distribution over a (finite) set of code words $Z$.

@@

The second challenge is that OCF requires the construction of a matrix whose rows and columns are indexed by the observed past and future sequences \citep{marzen-predictive-2016}.
Depending on the topological entropy of the process, the number of such sequences can grow as $|A|^M$, where $A$ is the set of observed symbols, and processes of interest often do show this exponential growth \citep{marzen-predictive-2016}.
Drastically, in the case of natural language, $A$ contains thousands of words.
We will see that this makes applying OCF infeasible even with $M=1$.

@@

\subsection{Variational Bound on Predictive Rate-Distortion}
We will first describe the general method, without committing to a specific framework for function approximation yet.
We will construct a bound on predictive rate-distortion and optimize this bound in a parametric family of function approximators to obtain an encoding $Z$ that is close to optimal for the nonparametric objective~(\ref{eq:ib}).

As in OCF (Section~\ref{sec:oce}), we assume that a set of finite sample trajectories $x_{-M} \dots x_{M-1}$ is available, and we aim to compress pasts of length $M$ to predict futures of length $M$. To carry this out, we restrict the PRD objective~(\ref{eq:ib}) to such finite-length pasts and futures: %\rljf{I think it would be a lot clearer to use the notation from other papers where it's like $\past$ except the arrow points to $M$. It's hard visually looking at the equations to distinguish $-M\dots-1$ from $0\dots M-1$, and that makes it harder to read. $^M\overleftarrow{X}$}
\begin{equation}\label{eq:ib-oce-repeat}
	\max_{Z : Z \bot \finitefuture | \finitepast} \left( \operatorname{I}[Z, \finitefuture] - \lambda \cdot \operatorname{I}[\finitepast, Z] \right).
\end{equation}
%where $Z$ ranges over all random variables such that the Markovcondition~(\ref{eq:markov}) holds:
%\begin{equation}
%    Z\bot \finitefuture | \finitepast
%\end{equation}
It will be convenient to equivalently rewrite (\ref{eq:ib-oce-repeat}) as
\begin{equation}\label{eq:prd-loss}
	\min_{Z : Z \bot \finitefuture | \finitepast}	\left[\operatorname{H}[\finitefuture | Z] + \lambda \cdot\operatorname{I}[ \finitepast, Z]\right],
\end{equation}
where $\operatorname{H}[\finitefuture | Z]$ is the \key{prediction loss}. Note that minimizing prediction loss is equivalent to maximizing predictiveness $\operatorname{I}[\finitefuture | Z]$. % is the reduction in prediction loss when predicting

@@

We now use the two propositions to rewrite the predictive rate-distortion objective~(\ref{eq:prd-loss}) in a way amenable to using function approximators.
Recall the predictive rate-distortion objective~(\ref{eq:prd-loss})
\begin{equation}\tag{\ref{eq:prd-loss}}
	\min_{Z : Z \bot \finitefuture | \finitepast}	\left[\operatorname{H}[\finitefuture | Z] + \lambda \operatorname{I}[ \finitepast, Z]\right]
\end{equation} % should be P_\psi (X_{0...M}|z)? % done (mhahn)
By the two Propositions, this objective is equivalent to
\begin{equation}\label{eq:bound}
	\min_{\phi, \psi, q}	\left[-	\mathbb{E}_{Z \sim \phi(\finitepast)}\left[\log P_\psi(\finitefuture | Z)\right] + \lambda \cdot \operatorname{D_{KL}}\infdivx{P_\phi(Z|\finitepast)}{q(Z)}\right],
\end{equation} % should be P_\psi (X_{0...M}|z)? % done (mhahn)
where $\phi, \psi, q$ range over all triples of the appropriate types described above.
From a solution to (\ref{eq:bound}), one obtains a solution to (\ref{eq:prd-loss}) by setting $Z$ to be the code generated by the encoder $\phi$.
The second proposition guarantees that, for a solution to  (\ref{eq:prd-loss}), the rate of this code is given as follows
\begin{equation}\label{eq:eq1-new}
 \operatorname{I}[Z, \finitepast] =   \mathbb{E}_{\finitepast}\left[ \operatorname{D_{KL}}\infdivx{P_\phi(Z|\finitepast)}{q(Z)}\right]
\end{equation}
and the prediction loss is given by
\begin{equation}\label{eq:eq2-new}
  %H[X_{0\dots M} | Z] =  -	\mathbb{E}_{z \sim \phi(\finitepast)}\left[\log P_\psi(X_{0\dots M} | z)\right]
  \operatorname{H}[\finitefuture|Z] =   - \mathbb{E}_{X_{-M\dots M-1}}	\mathbb{E}_{Z \sim P_\phi(\finitepast)}\left[\log P_\psi(\finitefuture | Z)\right]. %H[\finitefuture|\finitepast] % \operatorname{H}[\finitefuture]
\end{equation}
Note that the right-hand-sides of (\ref{eq:eq1-new}-\ref{eq:eq2-new}) can both be estimated efficiently using Monte Carlo samples from $Z \sim P_\phi(\finitepast)$.

@@

We consider the problem of estimating rate-distortion for natural language.
Natural language has been a testing ground for information-theoretic ideas since Shannon's work.
Much interest has been devoted to estimating the entropy rate of natural language~\citep{shannon1951prediction,takahira2016entropy,bentz2017entropy,takahashi2018cross}.
Indeed, the information density of language has been linked to human processing effort and to language structure.
The word-by-word information content has been shown to impact human processing effort as measured both by per-word reading times~\citep{hale-probabilistic-2001,levy-expectation-based-2008,smith-effect-2013} and by brain signals measured through EEG~\citep{DBLP:conf/acl/FrankOGV13, kuperberg2016we}.
Consequently, prediction is a fundamental component across theories of human language processing  \citep{kuperberg2016we}.
Relatedly, the Uniform Information Density and Constant Entropy Rate hypotheses~\citep{fenk1980konstanz, genzel2002entropy, jaeger2007speakers} state that languages order information in sentences and discourse so that the entropy rate stays approximately constant. % , ferrer2013constant

>>>>

% Reviewers:
% Susanne Still
% Lukasz Debowski
% Kumiko Tanaka-Ishii
% Noga Zaslavsky
% Alexander Alemi

% TODO mention sam bowman's language VAE

% potentially cite https://arxiv.org/pdf/1609.05191.pdf

@@

This motivates the study of models that try to extract those key features of past observations that are most relevant to predicting the future.
%This results in a tradeoff between the cost of representing past observations and predicting the future well.
%This tradeoff has been formalized and computationally studied in a range of prior studies.
A general information-theoretic framework for this problem is provided by Predictive Rate-Distortion~~\citep{still-bottleneck-2014,marzen-predictive-2016}, also known as the Past-Future Information Bottleneck~\citep{creutzig-past-future-2009}.
The Predictive Rate-Distortion tradeoff seeks to find an encoding of past observations that is maximally informative about future observations, while satisfying a bound on the amount of information that has to be stored.
More formally, this framework trades off prediction loss in the future, formalized as cross-entropy, with the cost of representing the past, formalized as the mutual information between the past observations and the compressed representations of the past.
Due to its information-theoretic nature, this framework is extremely general and applies to processes across vastly different domains.
It has been applied to linear dynamical systems \citep{creutzig-past-future-2009,amir2015past}, but is equally meaningful for discrete dynamical systems~\citep{marzen-predictive-2016}.
For biological systems that make predictions about their environment, this corresponds to placing an information-theoretic constraint on the computational cost used for conditioning actions on past observations~\citep{genewein2015bounded}.

@@

Optimal Causal Filtering (OCF, \citet{still-optimal-2010}) addresses the problem of estimating Predictive Rate-Distortion from a finite sample of observation trajectories.
It does so by constructing a matrix of observed frequencies of different pairs of past and future observations.
However, such a method faces a series of challenges~\citep{marzen-predictive-2016}.
One is the curse of dimensionality:
Modeling long dependencies requires storing an exponential number of observations, which quickly becomes intractable for current computation methods.
This exponential growth is particularly problematic when dealing with processes with a large state space.
For instance, the number of distinct words in a human language as found in large-scale text data easily exceeds $10^5$, making storing and manipulating counts of longer word sequences very challenging.
A second challenge is that of overfitting:
When deploying a predictive model constructed via OCF on new data to predict upcoming observations, such a model can only succeed when the past sequences occurred in the sample that OCF was applied to.
This is because OCF relies on counts of full past and future observation sequences; it does not generalize to unseen past sequences.

@@

In Section~\ref{sec:background}, we formally introduce predictive rate-distortion, and discuss related notions of predictive complexity.
In Section~\ref{sec:OCF}, we describe the prior method of Optimal Causel Filtering (OCF).
In Section~\ref{sec:nprd}, we describe our method NPRD.
In Section~\ref{sec:experiments}, we validate NPRD on processes whose predictive rate-distortion is analytically known, showing that it finds essentially optimal predictive models.
In Section~\ref{sec:language}, we apply NPRD to data from five languages, providing the first estimates of predictive rate-distortion of natural language.

@@

The main prior method for estimating predictive rate-distortion from data is Optimal Causal Filtering (OCF, \citet{still-optimal-2010}).
This method approximates predictive rate-distortion using two approximations:
First, it replaces semi-infinite pasts and futures with bounded-length contexts, i.e., pairs of finite past contexts $(\finitepast := X_{-M}\dots X_{-1})$ and future contexts $(\finitefuture := X_{0}\dots X_{M-1})$ of some finite length $M$.\footnote{It is not crucial that past and future contexts have the same lengths, and indeed \citet{still-optimal-2010} do not assume this. We do assume equal length throughout this paper for simplicity of our experiments, though nothing depends on this.}
The PRD objective (\ref{eq:ib}) then becomes (\ref{eq:ib-oce}), aiming to predict length-$M$ finite futures from summary codes $Z$ of length-$M$ finite pasts:
\begin{equation}\label{eq:ib-oce}
	\max_{Z : Z \bot \finitefuture | \finitepast} \left( \operatorname{I}[Z, \finitefuture] - \lambda \cdot \operatorname{I}[\finitepast, Z] \right).
\end{equation}
Second, OCF estimates information measures directly from the observed counts of $(\finitepast), (\finitefuture)$ using the plug-in estimator of mutual information. %\rljf{Should be \dots instead of commas?}
With such an estimator, the problem in (\ref{eq:ib-oce}) can be solved using a variant of the Blahut-Arimoto algorithm \citep{tishby-information-1999}, obtaining an encoder $P(Z|\finitepast)$ that maps each observed past sequence $\finitepast$ to a distribution over a (finite) set of code words $Z$.

@@

The second challenge is that OCF requires the construction of a matrix whose rows and columns are indexed by the observed past and future sequences \citep{marzen-predictive-2016}.
Depending on the topological entropy of the process, the number of such sequences can grow as $|A|^M$, where $A$ is the set of observed symbols, and processes of interest often do show this exponential growth \citep{marzen-predictive-2016}.
Drastically, in the case of natural language, $A$ contains thousands of words.
%We will see that this makes applying OCF to language data infeasible even with $M=1$.

@@

\subsection{Main Result: Variational Bound on Predictive Rate-Distortion}
We will first describe the general method, without committing to a specific framework for function approximation yet.
We will construct a bound on predictive rate-distortion and optimize this bound in a parametric family of function approximators to obtain an encoding $Z$ that is close to optimal for the nonparametric objective~(\ref{eq:ib}).

As in OCF (Section~\ref{sec:OCF}), we assume that a set of finite sample trajectories $x_{-M} \dots x_{M-1}$ is available, and we aim to compress pasts of length $M$ to predict futures of length $M$. To carry this out, we restrict the PRD objective~(\ref{eq:ib}) to such finite-length pasts and futures: %\rljf{I think it would be a lot clearer to use the notation from other papers where it's like $\past$ except the arrow points to $M$. It's hard visually looking at the equations to distinguish $-M\dots-1$ from $0\dots M-1$, and that makes it harder to read. $^M\overleftarrow{X}$}
\begin{equation}\label{eq:ib-oce-repeat}
	\max_{Z : Z \bot \finitefuture | \finitepast} \left( \operatorname{I}[Z, \finitefuture] - \lambda \cdot \operatorname{I}[\finitepast, Z] \right).
\end{equation}
%where $Z$ ranges over all random variables such that the Markovcondition~(\ref{eq:markov}) holds:
%\begin{equation}
%    Z\bot \finitefuture | \finitepast
%\end{equation}
It will be convenient to equivalently rewrite (\ref{eq:ib-oce-repeat}) as
\begin{equation}\label{eq:prd-loss}
	\min_{Z : Z \bot \finitefuture | \finitepast}	\left[\operatorname{H}[\finitefuture | Z] + \lambda \cdot\operatorname{I}[ \finitepast, Z]\right],
\end{equation}
where $\operatorname{H}[\finitefuture | Z]$ is the \key{prediction loss}. Note that minimizing prediction loss is equivalent to maximizing predictiveness $\operatorname{I}[\finitefuture | Z]$. % is the reduction in prediction loss when predicting

@@

We now use the two propositions to rewrite the predictive rate-distortion objective~(\ref{eq:prd-loss}) in a way amenable to using function approximators:

\begin{corollary}[Main Result]
The predictive rate-distortion objective~(\ref{eq:prd-loss})
\begin{equation}\tag{\ref{eq:prd-loss}}
	\min_{Z : Z \bot \finitefuture | \finitepast}	\left[\operatorname{H}[\finitefuture | Z] + \lambda \operatorname{I}[ \finitepast, Z]\right]
\end{equation} % should be P_\psi (X_{0...M}|z)? % done (mhahn)
is equivalent to
\begin{equation}\label{eq:bound}
	\min_{\phi, \psi, q}	\left[-	\mathbb{E}_{Z \sim \phi(\finitepast)}\left[\log P_\psi(\finitefuture | Z)\right] + \lambda \cdot \operatorname{D_{KL}} [ \infdivx{P_\phi(Z|\finitepast)}{q(Z)}\right],
\end{equation} % should be P_\psi (X_{0...M}|z)? % done (mhahn)
where $\phi, \psi, q$ range over all triples of the appropriate types described above.

From a solution to (\ref{eq:bound}), one obtains a solution to (\ref{eq:prd-loss}) by setting $Z$ to be the code generated by the encoder $\phi$.
%The second proposition guarantees that, for a solution to  (\ref{eq:prd-loss}), 
The rate of this code is given as follows
\begin{equation}\label{eq:eq1-new}
 \operatorname{I}[Z, \finitepast] =   \mathbb{E}_{\finitepast}\left[ \operatorname{D_{KL}}\infdivx{P_\phi(Z|\finitepast)}{q(Z)}\right]
\end{equation}
and the prediction loss is given by
\begin{equation}\label{eq:eq2-new}
  %H[X_{0\dots M} | Z] =  -	\mathbb{E}_{z \sim \phi(\finitepast)}\left[\log P_\psi(X_{0\dots M} | z)\right]
  \operatorname{H}[\finitefuture|Z] =   - \mathbb{E}_{X_{-M\dots M-1}}	\mathbb{E}_{Z \sim P_\phi(\finitepast)}\left[\log P_\psi(\finitefuture | Z)\right]. %H[\finitefuture|\finitepast] % \operatorname{H}[\finitefuture]
\end{equation}
\end{corollary}
\begin{proof}
TODO write
\end{proof}

Note that the right-hand-sides of (\ref{eq:eq1-new}-\ref{eq:eq2-new}) can both be estimated efficiently using Monte Carlo samples from $Z \sim P_\phi(\finitepast)$.

@@

We consider the problem of estimating rate-distortion for natural language.
Natural language has been a testing ground for information-theoretic ideas since Shannon's work.
Much interest has been devoted to estimating the entropy rate of natural language~\citep{shannon1951prediction,takahira2016entropy,bentz2017entropy,takahashi2018cross}.
Indeed, the information density of language has been linked to human processing effort and to language structure.
The word-by-word information content has been shown to impact human processing effort as measured both by per-word reading times~\citep{hale-probabilistic-2001,levy-expectation-based-2008,smith-effect-2013} and by brain signals measured through EEG~\citep{DBLP:conf/acl/FrankOGV13, kuperberg2016we}.
Consequently, prediction is a fundamental component across theories of human language processing  \citep{kuperberg2016we}.
Relatedly, the Uniform Information Density and Constant Entropy Rate hypotheses~\citep{fenk1980konstanz,genzel2002entropy,jaeger2007speakers} state that languages order information in sentences and discourse so that the entropy rate stays approximately constant. % , ferrer2013constant
